Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-18 ieee_paper.tex
description of the host can go out of bandwidth
performance section

@@

In this experimental section, as a point of reference we provide some
results on a typical CPU architecture. This system is ...\todo{include
  a short description of the host.}  All the codes are compiled with
icc xxx and -O3 optimization. The MIC codes use the IMCI instructions
while the CPU codes use AVX instructions.

@@

\subsection{Computations}
%----------------------
Scaling results as a function of the number of processors, on the MIC and on the host. The host calculations
are performed for reference only, to bring out the dominance of the MIC due to its advanced swizzling and channel 
operations, not available on the host. The performance on the host is 3x slower than on the MIC. The large cache
of the host alleviates somewhat the lower memory bandwidth, but this only reduces the cost of transfer {\tt x}. 
The cost memory cost for {\tt A} and {\tt col\_id} is unaffected by the larger cache size.

The performance on the MIC scales linearly with the number of threads. We compare three cases, each with and without bandwidth reduction using Reverse Cuthil-McKee (RCM). There are $96^3$ RBF nodes, leading to a matrix with 884,736 rows. The four source vectors occupy more than 14M bytes, which exceeds the aggregate L2 cache of 31 Mbytes (7.8 Mfloats.)

\todo{Is the host architecture described? Problem: I had to change hosts a few weeks ago, and it has different characteristics}. 
%----------------------

\begin{figure*}
  \centering
  \subfigure[Performance of $y=Ax$ on the MIC. Squares: base 1/1 case,
    solid circles: 4/4 case implemented in C++, solid triangles: 4/4
    case implemented with MIC vector instructions. $-O3$ compilation
    options. Each symbol has two cases: with (red) and without (blue) 
    Reverse Cuthill-McKee bandwidth reduction. The grid is $96^3$.]
    {\includegraphics[width=.40\linewidth]{figures/mic_performance_nb_threads.pdf}\label{fig:perf_mic}}
%
\hspace{.4in}
%
  \subfigure[Performance on the host on a sparse matrix with $64^3$ rows. Manual
    vectorization brings significant improvement there as
    well.]{\includegraphics[width=.40\linewidth]{figures/64x64x64.pdf}\label{fig:perf_host}}
  \caption{Vectorization matters in practice. Differences in BW
    translate in different performance.\todo{this two plots should ``correlate'' }}
\end{figure*}

\begin{figure*}[t]
  \centering
  \subfigure[Supercompact and Random]{\includegraphics[width=.48\linewidth]{figures/random_supercompact.pdf}}
  \subfigure[Compact and RBF-FD]{\includegraphics[width=.48\linewidth]{figures/rbf_compact.pdf}}

  %\subfigure[Supercompact]{\includegraphics[width=.48\linewidth]{figures/supercompact_max_perf.pdf}}
  %\subfigure[Compact]{\includegraphics[width=.48\linewidth]{figures/compact_max_perf.pdf}} \\
  %\subfigure[Random]{\includegraphics[width=.48\linewidth]{figures/random_max_perf.pdf}}
  %\subfigure[RBF 3D]{\includegraphics[width=.48\linewidth]{figures/rbf_max_perf_run1.pdf}}
  \caption{Performance of "best code" accross matrix types. The two horizontal lines depict worst and 
  best case performance (Figure fig:gflops-peak-perf). RCM (blue, $\blacktriangle$) and no RCM (red,$\blacktriangledown$).}
\end{figure*}

Derivatives in an RBFFD formulation are expressed as a sparse
matrix/vector multiplication (SpMV). Using the full 61 nodes of a MIC
processor, and all four threads of each node, we achieve lowly 18
Gflops when calculating a single vector of a single
function. Bandwidth reduction has minimal effect since the cache is
large enough to store both the derivative matrix and the solution
vector. We increase the number of computations per byte transferred by
calculating four different derivatives of four different
functions. With bandwidth reduction, we speed the calculation by a
factor of 8, and achieve upwards of 140 Gflops.

>>>>

In this experimental section, as a point of reference we also provide
some results on a typical CPU architecture. This system is
...\todo{include a short description of the host.}  All the codes are
compiled with icc xxx and -O3 optimization. The MIC codes use the IMCI
instructions while the CPU codes use AVX instructions.

@@

@@

\subsection{Computations}

\begin{figure*}[t]
  \centering
  \subfigure[Supercompact and Random]{\includegraphics[width=.48\linewidth]{figures/random_supercompact.pdf}\label{fig:perf_ideal}}
  \subfigure[Compact and RBF-FD]{\includegraphics[width=.48\linewidth]{figures/rbf_compact.pdf}\label{fig:perf_realistic}}

  \caption{Performance of "best code" accross matrix types. The two
    horizontal lines depict worst and best case performance
    (Figure~\ref{fig:gflops-peak-perf}). RCM (blue, $\blacktriangle$)
    and no RCM (red,$\blacktriangledown$).\todo{can we reorder the
      legend to give supercompact first in (a) and compact first in
      (b)?}}
  \label{fig:expe_types}
\end{figure*}

We now investigate the actual performance that we can obtain when
multiplying 4 vectors by 4 matrices in single
precision. Figure~\ref{fig:expe_types} present results for different
types of matrices and also shows the minimum and maximum performance
predicted in Section~\ref{sec:model}. Figure~\ref{fig:perf_ideal}
gives the results for the {\tt supercompact} and {\tt random} matrices
which represent both the best and the worst case for such a
computation. The {\tt supercompact} case peaks at
205Gflop/s\todo{check that} being very close to the predicted peak
performance of 213Gflop/s. Conversely the performance of the {\tt
  random} case decreases to 59Gflop/s\todo{check} which is close to
the lowest predicted performance of 55Gflop/s. One can see that RCM
ordering helps the {\tt random} case but the impact decreases when the
size of the matrix increase.

Figure~\ref{fig:perf_realistic} gives the results on the {\tt compact}
case which represents the best realistic matrix one could find and in
an {\tt rbf} matrix extracted from a real application. In the {\tt
  compact} case, the performance can be as high as 195Gflop/s being
within 15\% of the peak predicted performance. The performance of the
{\tt rbf} case varies between 100Gflop/s and 140Gflop/s. Most of the
time, the RCM ordering provide an improvement which can be as high as
30Gflop/s.

\begin{figure*}
  \centering
  \subfigure[Performance of $y=Ax$ on the MIC. Squares: base 1/1 case,
    solid circles: 4/4 case implemented in C++, solid triangles: 4/4
    case implemented with MIC vector instructions. $-O3$ compilation
    options. Each symbol has two cases: with (red) and without (blue) 
    Reverse Cuthill-McKee bandwidth reduction. The grid is $96^3$.]
    {\includegraphics[width=.40\linewidth]{figures/mic_performance_nb_threads.pdf}\label{fig:perf_mic}}
%
\hspace{.4in}
%
  \subfigure[Performance on the host on a sparse matrix with $64^3$ rows. Manual
    vectorization brings significant improvement there as
    well.]{\includegraphics[width=.40\linewidth]{figures/64x64x64.pdf}\label{fig:perf_host}}
  \caption{Vectorization matters in practice. Differences in BW
    translate in different performance.\todo{this two plots should ``correlate'' }}
  \label{fig:comparison_spmv}
\end{figure*}

We finally investigate the central questions of this paper. Do we gain
actual performance by transforming classical SpMV computation into the
multiplication of 4 vectors by 4 matrices for the computation of the
derivative of RBFs? Does using manual vectorization improve actual
application performance? Figure~\ref{fig:comparison_spmv} compares the
performance achieved by a classical SpMV and by the multiplication of
4 vectors by 4 matrices using either standard C++ code and our
optimized implementation. On MIC (Figure~\ref{fig:perf_mic}) the
classical SpMV computation reaches 15Gflop/s\todo{check}. The standard
C++ implementation reaches a performance of 40Gflop/s while our
optimized implementation reaches 137Gflop/s\todo{check}. Notice that
RCM ordering has no impact in the standard C++ implementation while it
provides an significant improvement in our manually vectorized
version. This highlights the C++ implementation is instruction bound
while our manually vectorized implementation is most likely memory
bound.  As a point of comparison, we show an AVX implementation on the
CPU of the two kernels and we show the best achievable performance in
the classical SpMV in Figure~\ref{fig:perf_host}. \todo{waiting for
  complete plot before writing more.}

%% Scaling results as a function of the number of processors, on the MIC
%% and on the host. The host calculations are performed for reference
%% only, to bring out the dominance of the MIC due to its advanced
%% swizzling and channel operations, not available on the host. The
%% performance on the host is 3x slower than on the MIC. The large cache
%% of the host alleviates somewhat the lower memory bandwidth, but this
%% only reduces the cost of transfer {\tt x}.  The cost memory cost for
%% {\tt A} and {\tt col\_id} is unaffected by the larger cache size.

%% The performance on the MIC scales linearly with the number of
%% threads. We compare three cases, each with and without bandwidth
%% reduction using Reverse Cuthil-McKee (RCM). There are $96^3$ RBF
%% nodes, leading to a matrix with 884,736 rows. The four source vectors
%% occupy more than 14M bytes, which exceeds the aggregate L2 cache of 31
%% Mbytes (7.8 Mfloats.)

%% \todo{Is the host architecture described? Problem: I had to change
%%   hosts a few weeks ago, and it has different characteristics}.
%% %----------------------

%% Derivatives in an RBFFD formulation are expressed as a sparse
%% matrix/vector multiplication (SpMV). Using the full 61 nodes of a MIC
%% processor, and all four threads of each node, we achieve lowly 18
%% Gflops when calculating a single vector of a single
%% function. Bandwidth reduction has minimal effect since the cache is
%% large enough to store both the derivative matrix and the solution
%% vector. We increase the number of computations per byte transferred by
%% calculating four different derivatives of four different
%% functions. With bandwidth reduction, we speed the calculation by a
%% factor of 8, and achieve upwards of 140 Gflops.
